Supplementary Information

### 2016-12-03 Response_to_reviewers.Rmd

We agree. Pathogen outbreaks cannot be ignored, because their impacts are significant for Ulmus, and (outside the current range) for Castanea, but they are not central to this paper. We have reduced their importance in the manuscript, but continue to acknowledge them as a factor in changing distributions in the Discussion and in the Introduction.  We’ve also added several references about the past effects of pathogens on tree mortality and forest composition.

@@

> Most importantly, the way that the authors included the effects of land use change in data analyses (let’s forget pathogen outbreaks for now) was not justified. Thus, I was not convinced the data analyses were robust to answer the question about the effects of climate change, land use change, and pathogen outbreaks on distribution changes. Although the authors argued the land use effects were represented as the difference between VHCH, VMCH. Why?

The reviewer is pointing out an important distinction:  what we can observe is historical climate change and vegetation change.  Yet we often make inferences about land use because, in this region, forest composition has been so heavily altered by past and current land use.   Previous authors have clearly indicated that EuroAmerican settlement in the north-central United States can be almost entirely attributed to human agency. Euro-American Settlement is associated with a suite of land use changes, including logging, associated slash fires, and large scale conversion of forested landscape to agriculture. Shulte et al (2007) and Rhemtulla et al (2009) have used similar language to explain the set of changes associated with EuroAmerican settlement. Initial clearance for agricultural and urban development was contemporaneous with massive logging operations, and, in the late 1800s, widespread fire originiating from human sources. Given this, and the fact that very little "old growth" forest remains, modern forests are indelibly imprinted with the signal of human land use, whether agricultural conversion or timbering. The IPCC considers forestry to be a component of land use. The Peshtigo firestorm, the Michigan fires of 1871 and 1881 were driven by slash buildup following logging. While sucession processes occurred since clearance and settlement, these were directly affected by the initial land clearance. For example, Rhemtulla et al. (2009) indicate that the loss of hemlock was caused both by logging, and recruitment failure as the result of the loss of seed source because of the extensive logging, thus both cause & effect are rooted in EuroAmerican settlement.

Thus, VH, from the PLSS forest survey (ignoring climate data for now), mostly represents forest structure and distribution in the absence of settlement impacts. VM represents a modern landscape, sampled over the last decade as part of the FIA program. Differences between VH and VM is almost entirely attributable to anthropogenic agency, either loss of forested land (oak-hickory forest lost 5x106 ha of coverage since settlement due to agricultural conversion, as pointed out in the response to Reviewer 2) or logging, tree plantations, or other forms of forestry use.  Hence we are confident in ascribing most changes from VC to VM to land use.
To further clarify, we have modified the explanation of Figure 2 in the Methods as follows:
*The effect of vegetation change on shifting PDFs is $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a). Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land-use change associated with Euro-American settlement.*

@@

> Firstly, it is not possible to get the distribution data under VMCH. Secondly, the distribution changes over the last two centuries were caused by many factors including land use change, disturbances (e.g., pathogen outbreaks, fire, harvest), and endogenous succession processes. Why the difference VH and VM could represent the land use change effects? This was too arbitrary. A stronger rationale is needed.

See response to prior comment

> The authors should use present tense not past tense in the methods and results sections.

In our experience, either present or past tense is acceptable, as long as verb tense is used consistently.  We have not changed tense, but are happy to comply with any request by Ecology Letters.

@@

> The results section was not well organized. Subtitles should be used to describe the overall distribution changes, shifts along climate change, the relative importance, and interactions. The four paragraphs of interaction subsection in the results section were not concise and like bullets.

We have ensured that each paragraph in the results has a clear and informative topic sentence.

@@

> Secondly, the materials and method section was vague and not logically articulated.
1) July temperature (Tmax) was the minimum, maximum, or mean temperature in July?
January temperature (Tmin) was the minimum, maximum, or mean temperature in January?
Annual temperature range (Tdiff) was the difference between the minimum and maximum temperature over the year? Annual precipitation (Pann) was the minimum, maximum, or mean precipitation across over the year?

We have clarified the description of these variables in the Methods: "*Estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) . . .*"

@@

> 2)  More detailed information in the materials and method section was needed to show how the authors analyzed the overall distribution changes, climate change effects, and the relative importance of land use and climate change. It should not be only in the figure titles. For example:

> The authors used gain, loss, and continuous persistence to demonstrate the distribution changes over the last two centuries in Figure 3. However, this approach was not described in the materials and method section but was only in the title of Figure 3.

We have added the following sentence to the Methods:

*In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "presence", where a tree was present within a grid cell in both FIA and PLS eras.*

@@

> The authors used the box-and-whisker plots to demonstrate the distribution changes along Tmax, Tmin, and Pann. Likewise, this approach was not described in the materials and method section but was only in the title of Figure 4.

We’ve amended a sentence in Methods to read “*Distributions of tree genera within climate space are shown using box plots and described as univariate probability density functions (PDFs) of tree species presence*”

@@

> The approach used in Figure 5 was not described in the materials and method section but was only in the title of Figure 5. More detailed information should be given to better analyze the relative importance of land use change and climate change effects.

The approach shown in Fig. 5 is described in the section of Methods ‘Shifts of Species Distributions in Climate Space.’   We have edited it for clarity.

@@

We have added the equation for the Hellinger distance:

*Hellinger distance for two discrete probability distributions is defined as:*

H(P, Q) = \frac{1}{\sqrt{2}} \; \sqrt{\sum_{i=1}^k (\sqrt{p_i} - \sqrt{q_i})^2}

*where $P$ and $Q$ are the distributions, with a common index $i$.*

@@

> 3) Again, how the effects land use and exotic pathogens were included in the analyses?

See responses to prior comments.

> 4) The 2×2 factorial design with pre-settlement and modern vegetation and climates (VHCH, VHCM,VMCH,VMCM) was very confusing (Figure 2a). The historical distribution data for VHCH scenario were derived from PLS and the contemporary distribution data for VMCM were derived from US FIA data. Then, what were the data for VHCM and VMCH scenarios. It was impossible to get the data for VHCM and VMCH scenarios, right? If my understanding was wrong, then the more detailed information was needed to better describe the factorial design and data used for each scenario.

We revised this section of the methods for clarity, in part by re-writing this paragraph:

*The total observed shift ($d_{tot}$) between a historical and modern PDF is the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and *vice versa*.  The effect of climate change on shifting PDFs ($d_{c}$) is the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). The effect of vegetation change on shifting PDFs is $d_{v}$, the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ (top row of Figure 2a). Because Euro-American land-use is the dominant driver of vegetation change in the upper Midwest over the past two centuries (Introduction), we often attribute shifts recorded by $d_{v}$ to the land-use change associated with Euro-American settlement.*

@@

> 5) About the t-tests to assess the interaction effects between land use and climate change (e.g., compounding, confounding, counteracting), what the responsible variables were used for the comparisons between the VHCH and VHCM, VHCH and VMCH. More detailed information was needed to better describe these tests.

We have modified the t-tests to include spatial autocorrelation and expanded their description in the Supplementary Information.

> Thirdly, the discussion section was not well written. Many contents in the discussion section ventured beyond the domain of the results. I suggest that the authors should set the results in a broader context and logically discuss the implications for distribution-climate relationships, distribution modeling, historical database, and management.

We have heavily revised the discussion in response to all reviewers’ comments.  Note that we were unsure about how to best address this comment, because it seems to ask that the discussion be both broadened and narrowed.  We hope that the revised discussion addresses these concerns.

> 1) L316-L322 were not the findings of this study. How were these points related to the findings? They read like introduction and should not be in discussion section.

This paragraph is intended to review mechanisms by which species-climate relationships may change over time. We have not changed this text.

@@

> 2) Likewise, L323-l336 how were these points related to your findings? It stood alone now.

We have added two pointers back to figure 1, to better link this discussion to that figure.

@@

> 1) P2-L36, change “is poorly understood” to “are poorly understood”.

Fixed.

> 2) P2-L41, change “implicates” to “implicated”.

This edit goes back to prior stylistic disagreements about choice of present vs. past tense.  We have kept the abstract in the present tense.

@@

> 4) The figures were not well made. The axis titles in Figure 1 were two small. The four lines in the Figure 2 b-e were difficult to read. The genera names in Figure 5 were too crowd and hard to tell. It may be better to use shorter names instead.

We have edited the figures to address these comments.

> 5) All figure titles were too long. Many of these contents were approach and should be put in methods section.

5)	We have added more information to the Methods section, while also retaining much of this information in the Figure legends.  We prefer a style that uses longer legends, to facilitate figure interpretation.

******************************************
